---
title: "Stochastic HPLC"
subtitle: "First-Passage Animation + Chromatogram of interconverting species"
format:
  dashboard:
    scrolling: true
    orientation: rows
    theme: cosmo
server: shiny
---

```{r}
#| context: setup
#| include: false
library(shiny)
library(dplyr)
library(tidyr)
library(ggplot2)
library(av)
library(ragg)
```

# Controls {.sidebar}

```{r}
actionButton("run", "Run simulation", class = "btn-primary")
sliderInput("N", "Particles (N)", min = 100, max = 1000, value = 500, step = 50)
sliderInput("L", "Column length", min = 10, max = 50, value = 20, step = 10)
sliderInput("v", "Flow rate / drift", min = 0.1, max = 1.0, value = 0.1, step = 0.1)
sliderInput("Dax", "Axial dispersion", min = 0.00, max = 0.10, value = 0.01, step = 0.01)
sliderInput("tau_M1", "Mean fly time (species 1)", min = 0.05, max = 1.00, value = 0.80, step = 0.05)
sliderInput("tau_S1", "Mean sojourn time (species 1)", min = 0.05, max = 1.00, value = 0.05, step = 0.05)
sliderInput("tau_M2", "Mean fly time (species 2)", min = 0.05, max = 1.00, value = 0.05, step = 0.05)
sliderInput("tau_S2", "Mean sojourn time (species 2)", min = 0.05, max = 1.00, value = 0.80, step = 0.05)
sliderInput("k12", "1→2 conversion rate constant", min = 0.00, max = 0.05, value = 0.00, step = 0.01)
sliderInput("k21", "2→1 conversion rate constant", min = 0.00, max = 0.05, value = 0.00, step = 0.01)
sliderInput("temp_scl", "Temporal resolution scale", min = 1, max = 100, value = 50, step = 1)
```

```{r}
#| context: server

simulate_hplc_fpt_2spec_interconv <- reactive({
  function(
    N, L, v, Dax,
    tau_M1, tau_S1,
    tau_M2, tau_S2,
    k12, k21,
    temp_scl) {
    
    set.seed(1)
    
    # column
    ID  <- 0.5
    eps <- 0.7
    S   <- pi * ID^2 / 4
    u   <- (v / S) / eps
    
    t0 <- L / u
    n  <- t0 / min(tau_M1, tau_M2)
    m2 <- 2 * n * max(tau_S1, tau_S2)^2
    dt <- round(sqrt(m2) / temp_scl, 3)
    
    # state (single pool)
    t      <- 0
    x      <- rep(0, N)
    alive  <- rep(TRUE, N)
    mobile <- rep(TRUE, N)
    fpt    <- rep(NA_real_, N)
    
    # species label: 1 or 2
    species <- rep(c(1L, 2L), each = N / 2)
    
    # per-step interconversion probabilities (from rates)
    p12 <- 1 - exp(-k12 * dt)
    p21 <- 1 - exp(-k21 * dt)
    
    # history (store both x and species for animate)
    x_hist       <- matrix(unname(x), nrow = 1)
    species_hist <- matrix(species, nrow = 1)
    
    n_alive_total <- N
    while (n_alive_total > 0) {
      t <- t + dt
      
      idx <- which(alive)
      if (!length(idx)) break
      
      # ---- (A) interconversion: 1 <-> 2 on alive particles ----
      sp <- species[idx]
      r  <- runif(length(idx))
      
      to2 <- which(sp == 1L & r < p12)
      if (length(to2)) species[idx[to2]] <- 2L
      
      to1 <- which(sp == 2L & r < p21)
      if (length(to1)) species[idx[to1]] <- 1L
      
      # ---- (B) update switching probabilities based on species ----
      # vectorized per-particle probs for alive particles
      p_off_mobile <- ifelse(species[idx] == 1L,
                             1 - exp(-dt / tau_M1),
                             1 - exp(-dt / tau_M2))
      
      p_off_stat   <- ifelse(species[idx] == 1L,
                             1 - exp(-dt / tau_S1),
                             1 - exp(-dt / tau_S2))
      
      # ---- (C) mobile <-> stationary switching ----
      m <- mobile[idx]
      
      r1 <- runif(length(idx))
      sw_m2s <- which(m & (r1 < p_off_mobile))
      if (length(sw_m2s)) mobile[idx[sw_m2s]] <- FALSE
      
      r2 <- runif(length(idx))
      sw_s2m <- which(!m & (r2 < p_off_stat))
      if (length(sw_s2m)) mobile[idx[sw_s2m]] <- TRUE
      
      # ---- (D) motion (recompute mobile after switching) ----
      idx_move <- idx[ mobile[idx] ]
      if (length(idx_move)) {
        x[idx_move] <- x[idx_move] + u * dt + sqrt(2 * Dax * dt) * rnorm(length(idx_move))
        x[idx_move] <- pmax(x[idx_move], 0)
      }
      
      # ---- (E) absorption ----
      hit <- idx[ x[idx] >= L ]
      if (length(hit)) {
        alive[hit] <- FALSE
        fpt[hit]   <- t
        x[hit]     <- L
        n_alive_total <- n_alive_total - length(hit)
      }
      
      # history
      x_hist       <- rbind(x_hist, unname(x))
      species_hist <- rbind(species_hist, species)
    }
    
    list(
      dt = dt,
      x_hist = x_hist,
      species_hist = species_hist,
      fpt = fpt
    )
  }
})

hist_to_df_interconv <- reactive({
  function(
    x_hist,
    species_hist,
    dt, N, L) {
    
    total_steps <- nrow(x_hist)
    frames <- total_steps
    
    frame_idx <- unique(round(seq(1, total_steps, length.out = frames)))
    frame_idx <- pmin(pmax(frame_idx, 1), total_steps)
    
    # fixed lane per particle (shared across species changes)
    y_lane <- runif(N, -0.45, 0.45)
    
    df <- tibble(
      frame = rep(seq_along(frame_idx), each = N),
      t     = rep((frame_idx - 1) * dt, each = N),
      id    = rep(seq_len(N), times = length(frame_idx)),
      x     = as.vector(t(x_hist[frame_idx, , drop = FALSE])),
      sp    = as.vector(t(species_hist[frame_idx, , drop = FALSE])),
      y     = rep(y_lane, times = length(frame_idx))
    ) %>%
      mutate(
        species  = if_else(sp == 1L, "sp1", "sp2"),
        absorbed = x >= (L - 1e-12)
      ) %>%
      select(-sp)
    
    df
  }
})

N        <- reactive({input$N})
L        <- reactive({input$L})
v        <- reactive({input$v})
Dax      <- reactive({input$Dax})
tau_M1   <- reactive({input$tau_M1})
tau_S1   <- reactive({input$tau_S1})
tau_M2   <- reactive({input$tau_M2})
tau_S2   <- reactive({input$tau_S2})
k12      <- reactive({input$k12})
k21      <- reactive({input$k21})
temp_scl <- reactive({input$temp_scl})

s <- eventReactive(input$run, {
  simulate_hplc_fpt_2spec_interconv()(
    N(), L(), v(), Dax(),
    tau_M1(), tau_S1(),
    tau_M2(), tau_S2(),
    k12(), k21(),
    temp_scl()
  )
})

dt           <- reactive({s()[["dt"]]})
x_hist       <- reactive({s()[["x_hist"]]})
species_hist <- reactive({s()[["species_hist"]]})
fpt          <- reactive({s()[["fpt"]]})

df <- reactive({
  hist_to_df_interconv()(
    x_hist(), species_hist(),
    dt(), N(), L()
  )
})

nframes <- reactive({max(df()$frame)})

df_fpt <- reactive({
  tibble(
    id  = seq_len(N()),
    fpt = fpt(),
    sp  = species_hist()[nrow(species_hist()), ]
  ) %>%
    filter(is.finite(fpt)) %>%
    mutate(species = if_else(sp == 1L, "sp1", "sp2")) %>%
    select(species, fpt)

})

x1     <- reactive({df_fpt()$fpt[df_fpt()$species == "sp1"]})
x2     <- reactive({df_fpt()$fpt[df_fpt()$species == "sp2"]})
x_all  <- reactive({c(x1(), x2())})
grid_n <- reactive({1024})
from   <- reactive({0})
to     <- reactive({max(x_all())})

bw_shared <- reactive({
  density(x_all(), n = grid_n(), from = from(), to = to())$bw
})

d1 <- reactive({
  density(x1(), bw = bw_shared(), n = grid_n(), from = from(), to = to(), adjust = 0.5)
})
d2 <- reactive({
  density(x2(), bw = bw_shared(), n = grid_n(), from = from(), to = to(), adjust = 0.5)
})

df_sum <- reactive({
  tibble(
    fpt      = d1()$x,
    dens_sum = d1()$y + d2()$y
  )
})

output$chromPlot <- renderPlot({
  df_fpt <- df_fpt()
  df_sum <- df_sum()
  ggplot(df_fpt, aes(x = fpt, colour = species, fill = species)) +
    geom_histogram(aes(y = after_stat(density)),
                   bins = 60, alpha = 0.20, position = "identity") +
    geom_line(data = df_sum, aes(x = fpt, y = dens_sum),
              inherit.aes = FALSE, colour = "black", linewidth = 1) +
    scale_fill_manual(values = c(
      sp1 = "#ff00cc",
      sp2 = "#339900"
    )) +
    scale_colour_manual(values = c(
      sp1 = "#ff00cc",
      sp2 = "#339900"
    )) +
    coord_cartesian(xlim = c(0, to()), expand = FALSE) +
    theme_minimal(base_size = 12) +
    labs(
      title = "Chromatogram proxy: species densities + kernel density estimate of sum",
      x = "First passage time",
      y = "Density"
    )
})

addResourcePath("media", tempdir())

mp4_path <- reactiveVal(NULL)

observeEvent(input$run, {
  req(df(), L())

  withProgress(message = "Rendering MP4", value = 0, {
    
    df0 <- df()
    L0  <- L()
    frames <- sort(unique(df0$frame))
    nF <- length(frames)

    frames_dir <- file.path(tempdir(), paste0("frames_", as.integer(runif(1, 1, 1e9))))
    dir.create(frames_dir, recursive = TRUE, showWarnings = FALSE)

    col_sp1 <- "#ff00cc"
    col_sp2 <- "#339900"
    outline <- "black"

    hex_alpha <- function(hex, alpha01) {
      rgbv <- grDevices::col2rgb(hex) / 255
      grDevices::rgb(rgbv[1], rgbv[2], rgbv[3], alpha = alpha01)
    }

    for (i in seq_along(frames)) {
      incProgress(0.8 / nF, detail = sprintf("Frame %d / %d", i, nF))

      f <- frames[i]
      d <- df0[df0$frame == f, , drop = FALSE]
      png_file <- file.path(frames_dir, sprintf("frame_%05d.png", i))

      if (requireNamespace("ragg", quietly = TRUE)) {
        ragg::agg_png(png_file, width = 900, height = 240, units = "px", res = 96)
      } else {
        grDevices::png(png_file, width = 900, height = 240)
      }

      op <- par(mar = c(0.5, 0.5, 2, 0.5))
      plot(NA, xlim = c(0, L0), ylim = c(-0.6, 0.6), axes = FALSE, xlab = "", ylab = "")
      rect(0, -0.55, L0, 0.55, col = hex_alpha("#add8e6", 0.15), border = NA)

      fill_base <- ifelse(d$species == "sp1", col_sp1, col_sp2)
      a <- ifelse(d$absorbed, 0.25, 0.9)
      fill <- mapply(hex_alpha, fill_base, a, SIMPLIFY = TRUE)

      points(d$x, d$y, pch = 21, bg = fill, col = outline, cex = 1.2, lwd = 1)
      title(main = "Particles in column")

      par(op)
      grDevices::dev.off()
    }

    incProgress(0.15, detail = "Encoding MP4…")
    mp4_file <- file.path(tempdir(), paste0("hplc_", as.integer(runif(1, 1, 1e9)), ".mp4"))
    pngs <- list.files(frames_dir, pattern = "\\.png$", full.names = TRUE)
    av::av_encode_video(pngs, output = mp4_file, framerate = 5)

    incProgress(0.05, detail = "Finalizing…")
    mp4_path(mp4_file)
  })
})

output$anim_ui <- renderUI({
  req(mp4_path())
  src <- file.path("media", basename(mp4_path()))
  tags$video(src = src, type = "video/mp4", controls = NA, autoplay = NA, loop = NA, width = "100%")
})
```

# Outputs

Parameters of the simulation can be set in the left sidebar, then press "Run simulation". A progress bar shows animation rendering.

Initial parameters set no interconversion. After setting the desired retention parameters, interconversion can be achieved by adjusting conversion rate constants.

Increase temporal resolution scale for a finer $\mathrm{d}t$ time grid during simulation of retention process.

## Animation

```{r}
#| title: Animation of retention and interconversion process
uiOutput("anim_ui")
```

## Smoothed FPT density (Chromatogram approximation)

```{r}
#| title: Histrogram of first passage times
plotOutput("chromPlot")
```
